---
title: "Pleiotropy in HPO/MGD/pan-UKB"
format: html
editor: visual
toc: true
toc-depth: 4
toc-indent: "4em"
toc-expand: true
---

## Environment

Here we're loading all of the necessary packages and objects for the bulk of the analysis:

```{r}
#| message: false
library(org.Hs.eg.db)
human_gotab <- select(org.Hs.eg.db, keys(org.Hs.eg.db, "GO"), "SYMBOL", "GO")
library(org.Mm.eg.db)
mouse_gotab <- select(org.Mm.eg.db, keys(org.Mm.eg.db, "GO"), "SYMBOL", "GO")

# Common part
library(ggplot2)
library(reshape2)
library(cowplot)
library(ggsci)
library(ggvenn)
library(RColorBrewer)
library(colorRamps)
library(clusterProfiler)
library(stringr)
library(msigdbr)
#library(org.Hs.eg.db)
#library(org.Mm.eg.db)

mypal = matlab.like2(10000)
my_ggrect <- element_rect(color='black', fill='#f5f5f5')
setwd("/mnt/data/work/systems_genetics/HPO_Pleiotropy/Pleiotropy paper/mbe_revision/pleiotropy_analysis/")
#load(".RData")
```

## Initial data import and export

The main dataset consists of a single dataframe with integrated data (produced during preliminary analysis steps).

```{r}
merged_data <- read.table('updated_general_HPO_table.tsv', 
                          sep='\t', header=T)
colnames(merged_data)
```

Besides, we export the GO term tables (for usage in another parts of the project).

```{r}
write.table(human_gotab, file='human_GO_data.tsv', 
            sep='\t', quote=F)
write.table(mouse_gotab, file='mouse_GO_data.tsv',
            sep='\t', quote=F)
```

## Distributions of the degree of pleiotropy

### SN - Analysis of various term counts

In this section, we will explore several key aspects of the PO term count distributions, answering the following questions:

-   Does the conversion of HPO terms to MP introduce substantial bias?

-   Do upper-level terms perform well in reducing spurious pleiotropy arising from correlated traits?

First, we create dataframes with the corresponding data, and plot the histograms of the term count per gene:

```{r}
merged_data$HP_high_count_noconvert <- sapply(merged_data$HP_high_level_human,
          function(x) length(strsplit(x, ',')[[1]]))
merged_data$HP_high_count_noconvert[merged_data$HP_high_count_noconvert == 0] = NA

# Terms vs clusters - SN section 1
human_counts <- merged_data[, c('HP_count', 'HP_high_count',
            'HP_high_count_noconvert', 'n_clusters_h')]
colnames(human_counts) = c('all_terms', 'mp_upper_terms', 
                           'hp_upper_terms', 'clusters')
human_counts <- na.omit(melt(human_counts))
human_counts$variable = factor(as.character(human_counts$variable),
                          levels=c('all_terms', 'clusters', 'hp_upper_terms',
                                   'mp_upper_terms'))
human_cnt_plot <- ggplot(human_counts, aes(x=value)) + 
  geom_histogram(binwidth = 1) +
  theme_bw() + facet_wrap(~variable, scales='free', nrow=4) +
  scale_x_continuous(limits=c(0, 100)) +
  xlab("Degree of pleiotropy") + ylab('Human gene count')
# human_cnt_plot

mouse_counts <- merged_data[, c('MP_count', 'MP_high_count', 'n_clusters_m')]
colnames(mouse_counts) = c('all_terms', 'upper_terms', 'clusters')
mouse_counts <- na.omit(melt(mouse_counts))
mouse_counts$variable = factor(as.character(mouse_counts$variable),
                levels=c('all_terms', 'clusters', 'upper_terms'))
mouse_cnt_plot <- ggplot(mouse_counts, aes(x=value)) + 
  geom_histogram(binwidth = 1) +
  theme_bw() + facet_wrap(~variable, scales='free', nrow=3) +
  scale_x_continuous(limits=c(0, 100)) +
  xlab("Degree of pleiotropy") + ylab('Mouse gene count')

sn1 <- plot_grid(human_cnt_plot, 
                 plot_grid(mouse_cnt_plot, NULL, rel_heights=c(1, 0.3), nrow=2),
                 nrow=1, labels=c('a', 'b'), label_size=20)
print(sn1)

```

Now we need some summary data regarding the numbers of non-pleiotropic genes as well as the median number of terms per gene.

```{r}
aggregate(value~variable, human_counts, function(x) sum(x == 1))
aggregate(value~variable, mouse_counts, function(x) sum(x == 1))
aggregate(value~variable, human_counts, median)
aggregate(value~variable, mouse_counts, median)
```

Now we check the overlap between sets of non-pleiotropic genes when using different ways of pleiotropy definition:

```{r}
human_clus_np <- na.omit(merged_data$gene_human[merged_data$n_clusters_h == 1])
human_upper_np <- na.omit(merged_data$gene_human[merged_data$HP_high_count == 1])
length(intersect(human_clus_np, human_upper_np))

mouse_clus_np <- na.omit(merged_data$gene_mouse[merged_data$n_clusters_m == 1])
mouse_upper_np <- na.omit(merged_data$gene_mouse[merged_data$MP_high_count == 1])
length(intersect(mouse_clus_np, mouse_upper_np))
```

Given the above data, we can conclude that clustering of terms does not seem beneficial compared to using raw lower-level terms or (especially) upper-level terms.

Now we have to return to the second question of HPO vs. MP, upper-level term counts. Below we will do two main things, i.e.: (i) look into the correlation of term counts, and (ii) look into the overlap of highly pleiotropic gene sets.

```{r}
ggplot(merged_data, aes(x=HP_high_count_noconvert, y=HP_high_count)) +
  geom_hex(bins=15, col='black', lwd=0.3) + theme_bw() +
  scale_fill_gradientn(colors=mypal, trans = "log",
                       breaks=c(1, 10, 100, 1000), name='Gene count') +
  theme(panel.grid=element_blank()) +
  xlab('HPO upper-level terms') +
  ylab('MP upper-level terms')
  
# Checking overlaps in highly pleiotropic genes
human_cut_hp <- quantile(na.omit(merged_data$HP_high_count_noconvert), 0.75)
human_cut_mp <- quantile(na.omit(merged_data$HP_high_count), 0.75)
hip_human_hp <- na.omit(merged_data$gene_human[merged_data$HP_high_count_noconvert > human_cut_hp])
hip_human_mp <- na.omit(merged_data$gene_human[merged_data$HP_high_count > human_cut_mp])
print(c(length(hip_human_hp), length(hip_human_mp), 
        length(intersect(hip_human_hp, hip_human_mp))))
```

### Final distributions for all sources

Here, we will plot the distributions of the degree of pleiotropy according to different datasets, and calculate some of the numbers describing the distributions.

Beginning with the distributions:

```{r}
# Figure 1a - distributions for selected term/GWAS vectors
pdeg_cols <- c('HP_high_count', 'MP_high_count',
               'pleio_degree')#, 'pleio_degree_locuswide')

pdeg_dist_data <- merged_data[, pdeg_cols]
source_names <- c('HPO', 'MGD', 'panUKB')#, 'panUKB_all')
names(source_names) = pdeg_cols
colnames(pdeg_dist_data) = source_names
pdeg_dist_data <- na.omit(melt(pdeg_dist_data))
pdeg_dist_data$Type = ifelse(grepl('panUKB', pdeg_dist_data$variable),
          'complex', 'Mendelian')
pdeg_distros <- ggplot(pdeg_dist_data, aes(x=value, fill=variable)) +
  geom_histogram(binwidth = 1, col='black', lwd=0.3) + theme_bw() +
  facet_wrap(~variable, nrow=1, scales='free') +
  scale_x_continuous(limits=c(0, 25)) +
  theme(panel.grid=element_blank(), strip.background=my_ggrect) +
  scale_fill_brewer(palette="Accent") +
  xlab('Degree of pleiotropy') + ylab('Gene count') +
  guides(fill=F)
pdeg_distros
aggregate(value~variable, pdeg_dist_data, median)
```

And, then, counts of associations per source, as well as some other metrics.

```{r}
assoc_cnt <- colSums(merged_data[, pdeg_cols], na.rm = T)
print(assoc_cnt)
sum(assoc_cnt)
gene_cnt <- colSums(merged_data[, pdeg_cols] > 0, na.rm = T)
print(gene_cnt)
non_pleio_or_missing <- function(x) {
  return(sapply(x, function(elt)
    ifelse(is.na(elt), TRUE, elt == 1)))
}

have_assoc_all <- sum(merged_data$HP_high_count > 0 & 
    merged_data$MP_high_count > 0 &
    merged_data$pleio_degree > 0, na.rm=T)
print(have_assoc_all)
have_assoc_both_dom <- sum((merged_data$HP_high_count > 0 | 
     merged_data$MP_high_count > 0) &
     merged_data$pleio_degree > 0, na.rm=T)
print(have_assoc_both_dom)
have_assoc_mono <- sum((merged_data$HP_high_count > 0 | 
       merged_data$MP_high_count > 0), na.rm=T)
print(have_assoc_mono)
pleio_mono <- sum(!non_pleio_or_missing(merged_data$HP_high_count) | !non_pleio_or_missing(merged_data$MP_high_count), na.rm=T)
print(pleio_mono)
# How many are pleiotropic in pan-UKB, but have no HPO/MGD
pleio_puk_notinmono <- sum(is.na(merged_data$HP_high_count) & 
    is.na(merged_data$MP_high_count) &
    merged_data$pleio_degree > 1, na.rm=T)
print(pleio_puk_notinmono)
# How many are uniquely pleiotropic in domains
pleio_mono_nopuk <- sum(
    (!non_pleio_or_missing(merged_data$HP_high_count) |
    !non_pleio_or_missing(merged_data$MP_high_count)) & 
    non_pleio_or_missing(merged_data$pleio_degree), na.rm=T)
print(pleio_mono_nopuk)
print(pleio_mono_nopuk/pleio_mono)
pleio_puk_nomono <- sum(
  (non_pleio_or_missing(merged_data$HP_high_count) &
  non_pleio_or_missing(merged_data$MP_high_count)) &
  !non_pleio_or_missing(merged_data$pleio_degree),
  na.rm=T
)
print(pleio_puk_nomono)
print(pleio_puk_nomono/sum(!non_pleio_or_missing(merged_data$pleio_degree)))
# Proportion of HPO and MGD pleiotropic that are not
# pleiotropic in pan-UKB
```

### Network modularity analysis

In this section, we will compute the modularity of the bipartite gene-trait networks using `igraph`. The initial implementation used `BRIM`, but the method has no stable implementation in either R or Python.

We will first run the analysis for pan-UKB (using the precomputed incidence matrix exported from Python), and then proceed to PO data.

```{r}
panukb_edges <- read.table('panUKB_closest_edgelist.txt',
                               sep='\t', header=F)
colnames(panukb_edges) = c('gene', 'cluster')

get_modularity <- function(edgelist_df) {
  incidence_matr <- table(edgelist_df)
  myg <- igraph::graph_from_biadjacency_matrix(incidence_matr,
                      mode=c("all"))
  myfc <- igraph::cluster_fast_greedy(myg)
  print(igraph::modularity(myfc))
}

get_modularity(panukb_edges)
```

Here, we will do the same for the ontology data, but first converting the information we have into an edge list.

```{r}
create_edgelist <- function(term_col='MP_high_level',
                             gene_col='gene_mouse',
                             full_df=merged_data) {
  df_snippet <- full_df[, c(term_col, gene_col)]
  get_all_nodes <- function(input_value, what_ret='term') {
    term_list = input_value[1]
    gene_name = input_value[2]
    if (is.na(term_list) || term_list == "") {
      return(c())
    }
    term_set = str_match_all(term_list, "MP:\\d+")[[1]][, 1]
    if (what_ret == 'term') {
      return(term_set)
    } else {
      return(rep(gene_name, length(term_set)))
    }
  }
  all_terms <- unlist(apply(df_snippet, 1, function(x)
    get_all_nodes(x, 'term'), simplify = T))
  all_genes <- unlist(apply(df_snippet, 1, function(x)
    get_all_nodes(x, 'gene'), simplify = T))
  edgelist_df <- data.frame(term=all_terms,
                            gene=all_genes)
  return(edgelist_df)
}
mgd_edgelist <- create_edgelist()
get_modularity(mgd_edgelist)
hpo_edgelist <- create_edgelist('HP_high_level', 'gene_human')
get_modularity(hpo_edgelist)

```

### SN - distributions for different ways of GWAS data processing

Below, we will draw two plots to compare the dgree of genic pleitoropy in pan-UKB data when using different ways of preprocessing the data (closest gene vs. all genes) or different sets of triats (clustered traits or max. independent trait set).

```{r}
sn2_data <- na.omit(melt(merged_data[, c('pleio_degree', 'pleio_degree_locuswide')]))
sn2_distros <- ggplot(sn2_data, aes(x=value, fill=variable)) +
  geom_histogram(binwidth = 1, col='black', lwd=0.3) + theme_bw() +
  facet_wrap(~variable, nrow=2, scales='free') +
  scale_x_continuous(limits=c(0, 25)) +
  theme(panel.grid=element_blank(), strip.background=my_ggrect) +
  scale_fill_brewer(palette="Set2") +
  xlab('Degree of pleiotropy') + ylab('Gene count') +
  guides(fill=F)
sn2_distros
# Counts
table(merged_data$pleio_degree > 0)
table(merged_data$pleio_degree == 1)/sum(merged_data$pleio_degree > 0, na.rm = T)
table(merged_data$pleio_degree_locuswide > 0)
table(merged_data$pleio_degree_locuswide == 1)/sum(merged_data$pleio_degree_locuswide > 0, na.rm = T)

# Comparison of distributions for max.indep for SN2.1
sn3_data <- na.omit(melt(merged_data[, c('pleio_degree', 'pleio_degree_maxindep')]))
sn3_distros <- ggplot(sn3_data, aes(x=value, fill=variable)) +
  geom_histogram(binwidth = 1, col='black', lwd=0.3) + theme_bw() +
  facet_wrap(~variable, nrow=2, scales='free') +
  scale_x_continuous(limits=c(0, 25)) +
  theme(panel.grid=element_blank(), strip.background=my_ggrect) +
  scale_fill_brewer(palette="Set3") +
  xlab('Degree of pleiotropy') + ylab('Gene count') +
  guides(fill=F)
sn3_distros
sum(merged_data$pleio_degree, na.rm=T)
sum(merged_data$pleio_degree_maxindep, na.rm=T)
mean(merged_data$pleio_degree, na.rm=T)
mean(merged_data$pleio_degree_maxindep, na.rm=T)
table(merged_data$pleio_degree > 1)
table(merged_data$pleio_degree_maxindep > 1)
cor.test(merged_data$pleio_degree, merged_data$pleio_degree_maxindep,
         use="complete.obs", method="spearman")
```

### Correlation analysis

Below, we construct a single plot to evaluate correlations between the genic degree of pleiotropy across our data.

```{r}
pdeg_scat_df <- data.frame(col_a=NA, col_b=NA, val_a=NA, val_b=NA)
pdeg_corrs <- data.frame(col_a=NA, col_b=NA, val_a=NA, val_b=NA, rval=NA, pval=NA)
for (i in pdeg_cols) {
  for (j in pdeg_cols) {
    if (i == j){
      next
    }
    df_snippet <- data.frame(col_a = source_names[i], 
                             col_b = source_names[j],
                             val_a = merged_data[, i],
                             val_b = merged_data[, j])
    pdeg_scat_df = na.omit(rbind(pdeg_scat_df, df_snippet))
    cor_test_res <- cor.test(merged_data[, i], merged_data[, j], 
                        use="complete.obs", method='spearman')
    pdeg_corrs <- na.omit(rbind(pdeg_corrs, 
                  data.frame(col_a = source_names[i], 
                             col_b = source_names[j],
                             val_a = 30, val_b = 30, 
                             rval = round(cor_test_res$estimate, digits=2),
                             pval = cor_test_res$p.value)))
  }
}
pdeg_scatters <- ggplot(pdeg_scat_df, aes(x=val_a, y=val_b)) +
  geom_hex(bins=15, col='black', lwd=0.3) + theme_bw() +
  geom_text(data=pdeg_corrs, aes(label=rval), size = 3) +
  facet_grid(vars(col_a), vars(col_b)) +
  scale_x_continuous(limits=c(0, 33)) + 
  scale_y_continuous(limits=c(0, 33)) +
  scale_fill_gradientn(colors=mypal, trans = "log",
        breaks=c(1, 10, 100, 1000), name='Gene count') +
  theme(panel.grid=element_blank(),
        strip.background = my_ggrect) +
  xlab('Degree of pleiotropy (dataset A)') +
  ylab('Degree of pleiotropy (dataset B)')
pdeg_scatters
# Adjusted p-values
p.adjust(unique(pdeg_corrs$pval), method='fdr')
```

### Overlap of pleiotropic gene sets

First, we calculate the generic overlap between the datasets.

```{r}
table(!(is.na(merged_data$pleio_degree_locuswide)) & 
        !(is.na(merged_data$HP_high_count)))
table(!(is.na(merged_data$pleio_degree)) & 
        !(is.na(merged_data$HP_high_count)))
```

Next, we collect sets of all pleiotropic genes for all datasets, including those analyzed for SN.

```{r}
pleio_genes_hpo <- na.omit(merged_data$gene_human[merged_data$HP_high_count > 1])
mouse_gene_4venn <- ifelse(is.na(merged_data$gene_human) |
        merged_data$gene_human == "",
        merged_data$gene_mouse, merged_data$gene_human)
pleio_genes_mgd <- na.omit(mouse_gene_4venn[merged_data$MP_high_count > 1])
pleio_genes_puk <- na.omit(merged_data$gene_human[merged_data$pleio_degree > 1])
pleio_genes_pukall <- na.omit(merged_data$gene_human[merged_data$pleio_degree_locuswide > 1])
pleio_genes_pukmi <- na.omit(merged_data$gene_human[merged_data$pleio_degree_maxindep > 1])

venn_pleio <- ggvenn(list('HPO' = pleio_genes_hpo, 'MGD' = pleio_genes_mgd,
            'panUKB' = pleio_genes_puk),#, 
#            'panUKB_all' = pleio_genes_pukall),
       set_name_size = 4, text_size=3, stroke_size=0.3,
       fill_color=brewer.pal(4, 'Accent'),
       show_percentage=FALSE)
venn_pleio

# Venns for SN2
sn2_venn_pleio <- ggvenn(list('closest' = pleio_genes_puk, 
                     'locus-wide' = pleio_genes_pukall),
                     set_name_size = 4, text_size=3, stroke_size=0.3,
                     fill_color=brewer.pal(2, 'Set2'),
                     show_percentage=FALSE)
sn2_venn_pleio
# Venn for SN3
sn3_venn_pleio <- ggvenn(list('clusters' = pleio_genes_puk, 
                              'max_indep_set' = pleio_genes_pukmi),
                         set_name_size = 4, text_size=3, stroke_size=0.3,
                         fill_color=brewer.pal(2, 'Set3'),
                         show_percentage=FALSE)
sn3_venn_pleio
```

Next, we turn to investigating the highly pleiotropic genes. We apply the 4th quartile rule to detect the highly pleiotropic genes irrespective of the actual distribution of the degrees of pleiotropy.

```{r}
hip_cut_hpo = quantile(na.omit(merged_data$HP_high_count), 0.75)
hip_cut_mgd = quantile(na.omit(merged_data$MP_high_count), 0.75)
hip_cut_puk = quantile(na.omit(merged_data$pleio_degree), 0.75)
hip_cut_pukall = quantile(na.omit(merged_data$pleio_degree_locuswide), 0.75)
hip_cut_pukmi = quantile(na.omit(merged_data$pleio_degree_maxindep), 0.75)

hip_genes_hpo <- na.omit(merged_data$gene_human[merged_data$HP_high_count > hip_cut_hpo])
hip_genes_mgd <- na.omit(mouse_gene_4venn[merged_data$MP_high_count > hip_cut_mgd])
hip_genes_puk <- na.omit(merged_data$gene_human[merged_data$pleio_degree > hip_cut_puk])
hip_genes_pukall <- na.omit(merged_data$gene_human[merged_data$pleio_degree_locuswide > hip_cut_pukall])
hip_genes_pukmi <- na.omit(merged_data$gene_human[merged_data$pleio_degree_maxindep > hip_cut_pukmi])

# Percentage of shared highly pleiotropic genes
hip_hpo_or_mgd <- union(hip_genes_hpo, hip_genes_mgd)
length(intersect(hip_hpo_or_mgd, hip_genes_puk))/length(hip_hpo_or_mgd)
```

And then make Venn diagrams for the highly pleiotropic genes similar to those for all pleiotropic genes.

```{r}
venn_hip <- ggvenn(list('HPO' = hip_genes_hpo, 'MGD' = hip_genes_mgd,
                          'panUKB' = hip_genes_puk), 
#                          'panUKB_all' = hip_genes_pukall),
                     set_name_size = 4, text_size=3, stroke_size=0.3,
                     fill_color=brewer.pal(4, 'Accent'),
                     show_percentage=FALSE)
venn_hip

# Highly pleiotropic gene venn diagram for SN2
sn2_venn_hip <- ggvenn(list('closest' = hip_genes_puk, 
                         'locus-wide' = hip_genes_pukall),
                         set_name_size = 4, text_size=3, stroke_size=0.3,
                         fill_color=brewer.pal(2, 'Set2'),
                         show_percentage=FALSE)
sn2_venn_hip
# Highly pleiotropic gene venn diagram for SN3
sn3_venn_hip <- ggvenn(list('clusters' = hip_genes_puk, 
                            'max_indep_set' = hip_genes_pukmi),
                       set_name_size = 4, text_size=3, stroke_size=0.3,
                       fill_color=brewer.pal(2, 'Set3'),
                       show_percentage=FALSE)
sn3_venn_hip
```

### Exporting figures with descriptive stats

In this chunk, we construct the plots that contain all panels for Figures 1, SN2.1, SN3.1. We hide the results of these chunks for ease of comprehension of the Qmd file.

```{r}
#| fig-show: hide
#| warning: false
#| message: false
# Main figure 1c-d
both_venn <- plot_grid(venn_pleio, venn_hip, labels=c('c', 'd'),
                       nrow=2, label_size = 20)
# Venns for Figure SN2.1 
sn2_venns <- plot_grid(sn2_venn_pleio, sn2_venn_hip,
                       nrow=1, labels=c('b', 'c'), label_size=20)
# FVenns for igure SN3.1 
sn3_venns <- plot_grid(sn3_venn_pleio, sn3_venn_hip,
                       nrow=1, labels=c('b', 'c'), label_size=20)

# Figure 1 --- 9 * 9
plot_grid(plot_grid(pdeg_distros, pdeg_scatters, nrow=2,
                    rel_heights = c(0.4, 1),
                    labels=c('a', 'b'), label_size=20),
          both_venn, nrow=1,
          rel_widths = c(1, 0.35), labels=c('', ''),
          label_size = 20)

# Figure SN2.1
plot_grid(sn2_distros, sn2_venns, nrow=2, labels=c('a', ''),
          label_size=20, rel_heights = c(1, 0.7))

# Figure SN3.1
plot_grid(sn3_distros, sn3_venns, nrow=2, labels=c('a', ''),
          label_size=20, rel_heights = c(1, 0.7))
```

## Comparison of gene features

### Necessary data manipulations

In this block, we will explore various features of genes depending on their degree of pleiotropy. To do so, we will need to first assign the groups of pleiotropy according to the 4th quartile rule described above.

```{r}
merged_data$pleiotropy_HPO = factor(ifelse(merged_data$HP_high_count > 1,
      ifelse(merged_data$HP_high_count > hip_cut_hpo, 'high', 'moderate'), 'none'),
      levels = c('none', 'moderate', 'high'))
merged_data$pleiotropy_MGD = factor(ifelse(merged_data$MP_high_count > 1,
      ifelse(merged_data$MP_high_count > hip_cut_mgd, 'high', 'moderate'), 'none'),
      levels = c('none', 'moderate', 'high'))
merged_data$pleiotropy_panUKB = factor(ifelse(merged_data$pleio_degree > 1,
      ifelse(merged_data$pleio_degree > hip_cut_puk, 'high', 'moderate'), 'none'),
      levels = c('none', 'moderate', 'high'))
merged_data$pleiotropy_panUKB_all = factor(ifelse(merged_data$pleio_degree_locuswide > 1,
      ifelse(merged_data$pleio_degree_locuswide > hip_cut_pukall, 'high', 'moderate'), 'none'),
      levels = c('none', 'moderate', 'high'))
merged_data$pleiotropy_panUKB_maxindep = factor(ifelse(merged_data$pleio_degree_maxindep > 1,
      ifelse(merged_data$pleio_degree > hip_cut_pukmi, 'high', 'moderate'), 'none'),
      levels = c('none', 'moderate', 'high'))
```

Next, we construct the plot of proportions of genes in different groups (for Extended Data Figure 1).

```{r}
group_props <- as.numeric(table(merged_data$pleiotropy_HPO))
group_props <- group_props/sum(group_props)
for (src_name in c(source_names[2:3])) {
  curr_props <- as.numeric(table(merged_data[, paste0('pleiotropy_', src_name)]))
  curr_props <- curr_props/sum(curr_props)
  group_props <- rbind(group_props, curr_props)
}
group_props <- as.data.frame(group_props)
colnames(group_props) <- c('none', 'moderate', 'high')
group_props$dataset <- source_names
# Extended Data Figure 1
ggplot(melt(group_props, id.vars='dataset'), aes(x=dataset, 
                      y=value, fill=variable)) +
  geom_bar(stat='identity', col='black') + theme_bw() +
  theme(panel.grid=element_blank(), 
        axis.text.x = element_text(angle=45, hjust=1)) +
  xlab('Data source') + ylab('Gene proportion') +
  labs(fill='Pleiotropy') +
  scale_fill_brewer(palette="BuPu")
```

Finally, we make a master function to plot all features we may want to, with the following imaportant args:

-   `axis_lab, ax_lim, log_scale` - regulating the axis properties on plots \[beware - axis limit is applied before plotting\];

-   `mouse_feature` - specification of a separate column for the value of the feature in mice (if this is different from the main feature column);

-   `gwas_data_file` - filename to use for GWAS locus-level data;

-   `cutoff, cut_name` - parameters for creating binary plots with a cutoff value to be used for splitting feature into categories;

-   `use_maxindep, use_locuswide` - arguments that, if set, override the set of columns used for analysis, juxtaposing the main (closest gene, clustered data) dataset against the selected one.

```{r}
plot_feature <- function(feature_name, axis_lab, ax_lim=NA, log_scale=FALSE,
                         mouse_feature=NA, gwas_data_file=NA, 
                         cutoff=NA, cut_name=NA, 
                         use_maxindep=FALSE, use_locuswide=FALSE) {
  if (use_maxindep) {
    source_names <- c('panUKB', 'panUKB_maxindep')
  } else if (use_locuswide) {
    source_names <- c('panUKB', 'panUKB_all')
  }
  pl_col_names <- paste0('pleiotropy_', source_names)
  feat_df <- merged_data[, c(feature_name, pl_col_names)]
  feat_df <- na.omit(melt(feat_df, id.vars=feature_name))
  colnames(feat_df) <- c('feature', 'dataset', 'group')
  feat_df$dataset = sapply(as.character(feat_df$dataset), 
                           function(x) strsplit(x, 'py_')[[1]][2])
  
  if (!is.na(mouse_feature)) {
    # Removing mouse data
    feat_df <- feat_df[feat_df$dataset != 'MGD', ]
    mouse_snippet <- na.omit(data.frame(feature=merged_data[, mouse_feature],
                                dataset='MGD',
                                group=merged_data$pleiotropy_MGD))
    feat_df <- as.data.frame(rbind(feat_df, mouse_snippet))
  }
  
  feat_df$group = factor(as.character(feat_df$group),
                         levels=c('none', 'moderate', 'high'))
  if (!is.na(gwas_data_file)) {
    gwas_values = read.table(gwas_data_file, sep='\t', header=T)
    feat_df <- gwas_values
    colnames(feat_df) <- c('group', 'feature')
    feat_df$group = factor(as.character(feat_df$group), 
                           levels=c('none', 'moderate', 'high'))
    feat_df$dataset <- 'panUKB (locus-wise)'
    source_names <- 'panUKB (locus-wise)'
  }
  
  summary_df <- aggregate(feature~group+dataset, feat_df, median)
  summary_df$pvalue = NA
  for (src_name in source_names) {
    for (tg_group in c('moderate', 'high')) {
      wtres <- wilcox.test(feature~group, feat_df[feat_df$dataset == src_name &
                  feat_df$group %in% c('none', tg_group), ])
      summary_df[summary_df$dataset == src_name  & 
                 summary_df$group == tg_group, 'pvalue'] = wtres$p.value
    }
  }
  summary_df$pvalue = p.adjust(summary_df$pvalue, method='fdr')
  print_sum_df <- summary_df
  
  if (is.na(ax_lim)) {
    ax_lim = max(feat_df$feature)
  } else {
    feat_df <- feat_df[feat_df$feature <= ax_lim, ]
  }
  if (log_scale) {
    line_shift <- ax_lim
  } else {
    line_shift = ax_lim/10
  }
  
  line_data <- data.frame(dataset=rep(source_names, each=4),
                  group=rep(c('none', 'moderate', 'none', 'high'), 
                            length(source_names)),
                  feature = rep(c(ax_lim + line_shift, ax_lim + line_shift, 
                      ax_lim + 2.5 * line_shift, ax_lim + 2.5 * line_shift), 
                      length(source_names)),
                  line_cat = rep(c(1, 1, 2, 2), length(source_names)))
  summary_df$plot_label = ifelse(summary_df$pvalue < 0.05,
                      ifelse(summary_df$pvalue < 0.01,
                      ifelse(summary_df$pvalue < 0.001, '***', '**'), '*'), 
                          'n.s.')
  summary_df$feature = rep(c(NA, ax_lim + 1.75 * line_shift, 
                           ax_lim + 3.25 * line_shift), length(source_names))
  
  # Shifting the asterisks a little
  summary_df$feature = ifelse(summary_df$plot_label == "n.s.",
        summary_df$feature + 0.05 * line_shift, 
        summary_df$feature - 0.25 * line_shift)

  out_plt <- ggplot(feat_df, aes(x=group, y=feature, fill=group)) +
    geom_violin(scale = "width") + theme_bw() + 
    theme(panel.grid = element_blank(), strip.background = my_ggrect) +
    geom_boxplot(outlier.shape=NA, width=0.25, col='black', fill='white') +
    geom_line(data=line_data, aes(group=line_cat)) +
    geom_text(data=summary_df[summary_df$group == 'moderate', ], 
              aes(label=plot_label), nudge_x=-0.5) +
    geom_text(data=summary_df[summary_df$group == 'high', ],
              aes(label=plot_label), nudge_x=-1) +
    ylab(axis_lab) + xlab('Pleiotropy group') +
    facet_wrap(~dataset, nrow=1) +  guides(fill=F) +
    scale_fill_brewer(palette='BuPu') +
    scale_y_continuous(limits=c(0, ax_lim * 1.4))
  
  if (log_scale) {
    out_plt <- out_plt + scale_y_log10()
  }
  
  if (!is.na(cutoff)) {
    cutoff_stats <- aggregate(feature~group+dataset, feat_df, function(x) 
      table(x > cutoff)/length(x))
    chisq_pvals <- c()
    for (src_name in source_names) {
      ds_df <- aggregate(feature~group, feat_df[feat_df$dataset == src_name & 
          feat_df$group %in% c('none', 'high'), ], function(x) table(x > cutoff))
      chisq_pvals <- c(chisq_pvals, chisq.test(ds_df$feature)$p.value)
    }
    cutoff_stats$no <- cutoff_stats$feature[, 1]
    cutoff_stats$yes <- cutoff_stats$feature[, 2]
    cutoff_stats <- cutoff_stats[, c(1, 2, 4, 5)]
    print(cutoff_stats)
    names(chisq_pvals) = source_names
    print(chisq_pvals)
    
    insert_asterisks <- function(this_src, 
                                 pvals=chisq_pvals) {
      asterisks <- ifelse(pvals[this_src] < 0.001,
            '***', ifelse(pvals[this_src] < 0.01,
            '**', ifelse(pvals[this_src] < 0.05, 
            '*', 'n.s.')))
      return(paste0(this_src, ' (', asterisks, ')'))
    }
    cutoff_df <- melt(cutoff_stats, measure.vars=c('no', 'yes'),
                      id.vars=c('group', 'dataset'))
    cutoff_df$dataset = sapply(cutoff_df$dataset,
                               insert_asterisks)
    prop_plot <- ggplot(cutoff_df, aes(x=group, y=value, fill=variable)) +
      geom_bar(stat='identity', col='black') + theme_bw() +
      xlab('Pleiotropy group') + ylab('Proportion of genes') +
      facet_wrap(~dataset, ncol=4) +
#      coord_flip() + 
#      facet_wrap(~dataset, nrow=4) +
      scale_fill_brewer(palette="OrRd", name=cut_name) +
      theme(strip.background = my_ggrect, 
            panel.grid = element_blank(),
            legend.position = 'top',
            axis.text.x=element_text(angle=45, hjust=1))
    return(prop_plot)
    
  } else {
    print(print_sum_df)
    return(out_plt)
  }
  
}
```

### Expression level analysis

Here, we work with several major summary statistics of the expression data according to the GTEx data: (i) mean or median expression across tissues; (ii) maximum expression across tissues; and (iii) number of tissues with detectable (\> 5 TPM) expression.

```{r}
count_plot <- plot_feature('count_exp', 'Number of tissues')
print(count_plot)

# Additional plots for supplementary information
median_plot <- plot_feature('median_exp', 'Median TPM', 100)
print(median_plot)
mean_plot <- plot_feature('mean_exp', 'Mean TPM', 100)
print(mean_plot)
max_plot <- plot_feature('max_exp', 'Max. TPM across tissues', 500)
print(max_plot)
```

```{r}
#| fig-show: hide
# Extended Data Figure 2
plot_grid(median_plot, mean_plot, max_plot, nrow=3, labels=c('a', 'b', 'c'),
          label_size=20)
```

### PPI and dosage sensitivity measures

In this section, we plot BIOGRID PPI counts and dosage sensitivity values from Collins *et al*., 2022.

```{r}
# PPI count
ppi_plot <- plot_feature('n_interactions', 'PPI count', ax_lim = 200)
print(ppi_plot)

# Collin et al., 2022 - pHaplo and pTriplo
phaplo_plot <- plot_feature('pHaplo', 'pHaplo')
print(phaplo_plot)
# Extended Data Figure 4
ptriplo_plot <- plot_feature('pTriplo', 'pTriplo')
print(ptriplo_plot)
```

### SN - plots of expression and PPI counts

Here, we plot the same features (number of tissues and the number of protein-protein interactors) for the max-indep and locus-wide pan-UKB data. These plots then go to SN2.2 and SN3.2

```{r}
count_plot_mi <- plot_feature('count_exp', 'Number of tissues', 
                              use_maxindep = TRUE)
print(count_plot_mi)
count_plot_lw <- plot_feature('count_exp', 'Number of tissues', 
                              use_locuswide =  TRUE)
print(count_plot_lw)

ppi_plot_mi <- plot_feature('n_interactions', 'PPI count', 
                              use_maxindep = TRUE, ax_lim=200)
print(ppi_plot_mi)
ppi_plot_lw <- plot_feature('n_interactions', 'PPI count', 
                            use_locuswide = TRUE, ax_lim=200)
print(ppi_plot_lw)
```

### GO term counts per gene

In this section, we make plots with GO term counts. To construct such plots, we define a funciton for annotating genes with term counts, and then run the annotation prior to plotting the results.

```{r}
add_count <- function(column_name, gotab_df, ont_name, gene_col='gene_human')  {
  cnt_vec <- table(gotab_df[gotab_df$ONTOLOGY == ont_name, 'SYMBOL'])
  merged_data[, column_name] = sapply(merged_data[, gene_col], function(x)
    cnt_vec[x])
  return(merged_data)
}

merged_data <- add_count('gobp_cnt_human', human_gotab, 'BP')
merged_data <- add_count('gomf_cnt_human', human_gotab, 'MF')
merged_data <- add_count('gocc_cnt_human', human_gotab, 'CC')
merged_data <- add_count('gobp_cnt_mouse', mouse_gotab, 'BP', gene_col='gene_mouse')
merged_data <- add_count('gomf_cnt_mouse', mouse_gotab, 'MF', gene_col='gene_mouse')
merged_data <- add_count('gocc_cnt_mouse', mouse_gotab, 'CC', gene_col='gene_mouse')

gobp_plot <- plot_feature('gobp_cnt_human', 'GO BP count', ax_lim=50,
                          mouse_feature = 'gobp_cnt_mouse')
print(gobp_plot)
gomf_plot <- plot_feature('gomf_cnt_human', 'GO MF count', ax_lim=50,
                          mouse_feature = 'gomf_cnt_mouse')
print(gomf_plot)
gocc_plot <- plot_feature('gocc_cnt_human', 'GO CC count', ax_lim=50,
                          mouse_feature = 'gocc_cnt_mouse')
print(gocc_plot)
```

Besides, we plot the GO biological process term counts for other variants of the pan-UKB GWAS data (for SN2/3 figures):

```{r}
gobp_plot_mi <- plot_feature('gobp_cnt_human', 'GO BP count', 
                            use_maxindep = TRUE, ax_lim=50)
print(gobp_plot_mi)
gobp_plot_lw <- plot_feature('gobp_cnt_human', 'GO BP count', 
                             use_locuswide = TRUE, ax_lim=50)
print(gobp_plot_lw)
```

And a plot for Extended Data (hidden from Qmd output).

```{r}
#| fig-show: hide
# Extended Data Figure 3
plot_grid(gocc_plot, gomf_plot, nrow=2, labels=c('a', 'b'))
# Figure 2 -- 8 * 7 (portrait)
plot_grid(count_plot, ppi_plot, gobp_plot, phaplo_plot,
          ncol=1, labels=c('a', 'b', 'c', 'd'), label_size = 20)
```

## Gene set enrichment analysis

### Basic enrichment testing

Here, we run the enrichment using the `clusterProfiler` package, without setting the universe (i.e., running against all genes in the genome). We run the analysis using only highly-pleiotropic gene subset.

First, we define function for running the enrichment:

```{r}
# Conveniences for EA
extract_dbname <- function(x, gs_sizes) {
  gs_size = gs_sizes[x]
  db_name <- str_match(x, 'KEGG|REACTOME|PID|BIOCARTA|WP|HALLMARK')[, 1]
  gs_name <- tolower(substr(x, str_length(db_name) + 2, 
                            str_length(x)))
  gs_name <- gsub('_', ' ', gs_name)
  gs_size_str <- paste0('(n=', gs_size, ')')
  return(paste(db_name, gs_name, gs_size_str, sep=' '))
}
make_enrichment_plot <- function(gene_list, orgdb_name,
                                 ontology, plot_title,
                                 data_source="PO",
                                 use_t2g=FALSE,
                                 t2g_df=NA,
                                 gs_sizes=NA) {
  if (use_t2g) {
    t2g_df$gs_name = sapply(t2g_df$gs_name, function(x)
                extract_dbname(x, gs_sizes))
    ego <- enricher(gene = gene_list, 
                    TERM2GENE = t2g_df,
                    pvalueCutoff = 0.05,
                    pAdjustMethod = "BH",
                    maxGSSize = 2500)
  } else {
    ego <- enrichGO(gene = gene_list,
                  OrgDb = orgdb_name,
                  keyType = "SYMBOL",
                  ont = ontology,
                  pvalueCutoff = 0.05,
                  pAdjustMethod = "BH",
                  maxGSSize = 2500,
                  readable = TRUE)
  }
  
  go_res_plot <- dotplot(ego, showCategory=10, title=plot_title)
  write.csv(ego, file = paste0(data_source, '_', orgdb_name, "_", ontology, ".csv"), 
            row.names = FALSE)
  return(list('df' = as.data.frame(ego), 'plot' = go_res_plot))
}
```

And then apply the function to gene sets:

```{r}
# Human
human_bp <- make_enrichment_plot(hip_genes_hpo, "org.Hs.eg.db", 'BP', 
                     'Biological process (human)')
print(human_bp$plot)
human_mf <- make_enrichment_plot(hip_genes_hpo, "org.Hs.eg.db", 'MF', 
                                 'Molecular functoin (human)')
print(human_mf$plot)
human_cc <- make_enrichment_plot(hip_genes_hpo, "org.Hs.eg.db", 'CC', 
                                 'Cellular component (human)')
print(human_cc$plot)

# Molecular function enrichment with all HPO genes
all_hpo_genes <- merged_data[!is.na(merged_data$pleiotropy_HPO),
                             'gene_human']
all_hpo_mf <- make_enrichment_plot(all_hpo_genes, 'org.Hs.eg.db', 'MF',
                              'Molecular function (all HPO genes')
print(all_hpo_mf$plot)
all_hpo_tf <- all_hpo_mf$df[grepl('transcription factor', 
                                  all_hpo_mf$df$Description), ]
hip_hpo_tf <- human_mf$df[grepl('transcription factor', 
                                human_mf$df$Description), ]
# Checking using 'DNA binding
all_hpo_dnab <- all_hpo_mf$df[grepl('DNA binding', 
                                  all_hpo_mf$df$Description), ]
hip_hpo_dnab <- human_mf$df[grepl('DNA binding', 
                                    human_mf$df$Description), ]


# Mouse
hip_genes_mgd_mm <- na.omit(merged_data$gene_mouse[merged_data$MP_high_count > hip_cut_mgd])
mouse_bp <- make_enrichment_plot(hip_genes_mgd_mm, "org.Mm.eg.db", 'BP', 
                                 'Biological process (mouse)')
print(mouse_bp$plot)
mouse_mf <- make_enrichment_plot(hip_genes_mgd_mm, "org.Mm.eg.db", 'MF', 
                                 'Molecular functoin (mouse)')
print(mouse_mf$plot)
mouse_cc <- make_enrichment_plot(hip_genes_mgd_mm, "org.Mm.eg.db", 'CC', 
                                 'Cellular component (mouse))')
print(mouse_cc$plot)
```

The same for pan-UKB data (we separate it from the rest because we should bear in mind that GSEA in GWAS may be biased by linkage disequilibrium).

```{r}
# panUKB analysis - undestandably flawed, but still important
puk_bp <- make_enrichment_plot(hip_genes_puk, "org.Hs.eg.db", 'BP', 
                      'Biological process (human)', data_source = "GWAS")
print(puk_bp$plot)
puk_mf <- make_enrichment_plot(hip_genes_puk, "org.Hs.eg.db", 'MF', 
                      'Molecular functoin (human)', data_source = "GWAS")
print(puk_mf$plot)
puk_cc <- make_enrichment_plot(hip_genes_puk, "org.Hs.eg.db", 'CC', 
                      'Cellular component (human)', data_source = "GWAS")
print(puk_cc$plot)
```

```{r}
#| fig-show: hide
# Extended Data Figure 7-9 (8 * 10 each)
plot_grid(human_bp$plot, mouse_bp$plot, puk_bp$plot,
          nrow=3, labels=c('a', 'b', 'c'), label_size = 20)
plot_grid(human_mf$plot, mouse_mf$plot, puk_mf$plot,
          nrow=3, labels=c('a', 'b', 'c'), label_size = 20)
plot_grid(human_cc$plot, mouse_cc$plot, puk_cc$plot,
          nrow=3, labels=c('a', 'b', 'c'), label_size = 20)
```

### Hallmark and canonical pathway analysis

#### Data retrieval and filtering

In this section, we endeavor to run the same EA framework but using the canonical pathway gene sets. We will run it in the two following ways:

-   Using hallmark (H) gene sets

-   Using the canonical pathway (CP) dataset;

-   Constructing the maximum independent set of pathways in an iterative algorithm

First, let's get the data from MSigDB using the `msigdbr`functions:

```{r}
# Hallmarks
hallmark_human_t2g = msigdbr(species = "human", 
                             category = "H") %>% 
       dplyr::distinct(gs_name, gene_symbol) %>% 
       as.data.frame()
hallmark_mouse_t2g = msigdbr(species = "mouse", 
                             category = "H") %>% 
       dplyr::distinct(gs_name, gene_symbol) %>% 
       as.data.frame()

cp_human_t2g = msigdbr(species = "human", category = "C2") %>% 
       dplyr::filter(gs_subcat != 'CGP') %>%
       dplyr::distinct(gs_name, gene_symbol) %>% 
       as.data.frame()
cp_mouse_t2g = msigdbr(species = "mouse", category = "C2") %>% 
       dplyr::filter(gs_subcat != 'CGP') %>%
       dplyr::distinct(gs_name, gene_symbol) %>% 
       as.data.frame()

gs_sizes <- c(table(cp_human_t2g$gs_name),
              table(hallmark_human_t2g$gs_name))
```

#### Raw H and CP testing

Now, we run the plain EA using the hallmarks:

```{r}
human_hall <- make_enrichment_plot(hip_genes_hpo,  "MSigDB",
                     "H", 'Hallmarks (human)', use_t2g = TRUE,
                     t2g_df=hallmark_human_t2g,
                     gs_sizes=gs_sizes)
print(human_hall$plot)
mouse_hall <- make_enrichment_plot(hip_genes_mgd_mm, "MSigDB",
                     "H", 'Hallmarks (mouse)', use_t2g = TRUE,
                     t2g_df=hallmark_mouse_t2g,
                     gs_sizes=gs_sizes)
print(mouse_hall$plot)
puk_hall <- make_enrichment_plot(hip_genes_puk, "MSigDB",
                     "H", 'Hallmarks (GWAS)', use_t2g = TRUE,
                     t2g_df=hallmark_human_t2g,
                     data_source = "GWAS",
                     gs_sizes=gs_sizes)
print(puk_hall$plot)
```

There are virtually no overlapping enriched gene sets, except for the `HALLMARK UV RESPONSE DN`. The overall number of enriched sets is below 10 for all sources except MGD. Hence, let's proceed to the CP.

```{r}
human_cp <- make_enrichment_plot(hip_genes_hpo,  "MSigDB",
              "CP", 'Canonical pathways (human)', use_t2g = TRUE,
              t2g_df=cp_human_t2g, gs_sizes=gs_sizes)
print(human_cp$plot)
mouse_cp <- make_enrichment_plot(hip_genes_mgd_mm, "MSigDB",
              "CP", 'Canonical pathways (mouse)', use_t2g = TRUE,
              t2g_df=cp_mouse_t2g, gs_sizes=gs_sizes)
print(mouse_cp$plot)
puk_cp <- make_enrichment_plot(hip_genes_puk, "MSigDB",
              "CP", 'Canonical pathways (GWAS)', use_t2g = TRUE,
              t2g_df=cp_human_t2g, gs_sizes=gs_sizes,
              data_source = "GWAS")
print(puk_cp$plot)
```

Some additional code for the figure export:

```{r}
#| fig-show: hide
# Extended Data Figure 5
plot_grid(human_hall$plot, mouse_hall$plot, puk_hall$plot,
          nrow=3, labels=c('a', 'b', 'c'), label_size=22)
# Extended Data Figure 6ss
plot_grid(human_cp$plot, mouse_cp$plot, puk_cp$plot,
          nrow=3, labels=c('a', 'b', 'c'), label_size=22)
```

#### Independent large pathways

Now let's try constructing the maximum independent set of CP pathways. To do so, we will do the following:

1.  Compute the number of genes per pathway;
2.  Take the largest one by the total gene count;
3.  Subtract genes belonging to the pathway in (2), return to step (1).
4.  Repeat until a certain number or fraction of genes are covered

```{r}
max_indep_gs <- function(t2g_df, gene_cutoff=15000,
                         increment_cutoff=50) {
  next_iteration = TRUE
  indep_gs_set <- c()
  seen_genes <- c()
  init_gs_gene_counts <- table(t2g_df$gs_name)
  while (next_iteration) {
    gs_gene_counts = table(t2g_df$gs_name)
    largest_gs <- names(gs_gene_counts)[which.max(gs_gene_counts)]
    largest_gs_genes <- t2g_df[t2g_df$gs_name == largest_gs, 
                               'gene_symbol']
    seen_genes <- c(seen_genes, largest_gs_genes)
    indep_gs_set <- c(indep_gs_set, largest_gs)
    # Logging
    print(noquote(paste(length(indep_gs_set),
          largest_gs, init_gs_gene_counts[largest_gs],
          length(largest_gs_genes),
          length(seen_genes), sep=' ')))
    t2g_df <- t2g_df[!(t2g_df$gene_symbol %in% largest_gs_genes), ]
    if (length(seen_genes) >= gene_cutoff || 
        nrow(t2g_df) == 0 ||
        length(largest_gs_genes) < increment_cutoff) {
      break
    }
  }
  return(indep_gs_set)
}

indep_cp_human <- max_indep_gs(cp_human_t2g,
                               increment_cutoff = 25)
cp_indep_human_t2g <- cp_human_t2g[cp_human_t2g$gs_name %in%
                                     indep_cp_human, ]
cp_indep_mouse_t2g <- cp_mouse_t2g[cp_mouse_t2g$gs_name %in%
                                     indep_cp_human, ]
```

Following the construction of this set of pathways, we now run the EA against it:

```{r}
human_indepcp <- make_enrichment_plot(hip_genes_hpo,  "MSigDB",
        "CPindep", 'CP-independent (human)', use_t2g = TRUE,
        t2g_df=cp_indep_human_t2g, gs_sizes=gs_sizes)
print(human_indepcp$plot)
mouse_indepcp <- make_enrichment_plot(hip_genes_mgd_mm, "MSigDB",
        "CPindep", 'CP-independent (mouse)', use_t2g = TRUE,
        t2g_df=cp_indep_mouse_t2g, gs_sizes=gs_sizes)
print(mouse_indepcp$plot)
puk_indepcp <- make_enrichment_plot(hip_genes_puk, "MSigDB",
        "CPindep", 'CP-independent (GWAS)', use_t2g = TRUE,
        t2g_df=cp_indep_human_t2g, gs_sizes=gs_sizes,
        data_source = "GWAS")
print(puk_indepcp$plot)
```

We can also calculate the proportion of gene sets out of the constructed list that have a significant enrichment, and find out the common ones.

```{r}
nrow(human_indepcp$df)
nrow(mouse_indepcp$df)
nrow(puk_indepcp$df)
intersect(intersect(human_indepcp$df$ID, mouse_indepcp$df$ID),
          puk_indepcp$df$ID)
```

### Evaluating overlap of enriched terms

Next, we use the results of GSEA obtained above to evaluate the overlap between the sets of enriched terms (in each of the GO branches).

```{r}
# Analysis of similarity of results between data types (HPO vs. pan-UKB)
common_bp <- intersect(human_bp$df$ID, puk_bp$df$ID)
length(common_bp)
length(common_bp)/min(nrow(human_bp$df), nrow(puk_bp$df))

common_mf <- intersect(human_mf$df$ID, puk_mf$df$ID)
length(common_mf)
length(common_mf)/min(nrow(human_mf$df), nrow(puk_mf$df))

common_cc <- intersect(human_cc$df$ID, puk_cc$df$ID)
length(common_cc)
length(common_cc)/min(nrow(human_cc$df), nrow(puk_cc$df))

venn_terms_bp <- ggvenn(list('HPO' = human_bp$df$ID, 
                             'MGD' = mouse_bp$df$ID,
                        'panUKB' = puk_bp$df$ID),
                   set_name_size = 4, text_size=3, stroke_size=0.3,
                   fill_color=brewer.pal(3, 'Accent'),
                   show_percentage=FALSE)
venn_terms_cc <- ggvenn(list('HPO' = human_cc$df$ID, 
                             'MGD' = mouse_cc$df$ID,
                        'panUKB' = puk_cc$df$ID),
                   set_name_size = 4, text_size=3, stroke_size=0.3,
                   fill_color=brewer.pal(3, 'Accent'),
                   show_percentage=FALSE)
venn_terms_mf <- ggvenn(list('HPO' = human_mf$df$ID, 
                             'MGD' = mouse_mf$df$ID,
                        'panUKB' = puk_mf$df$ID),
                   set_name_size = 4, text_size=3, stroke_size=0.3,
                   fill_color=brewer.pal(3, 'Accent'),
                   show_percentage=FALSE)

# 6 * 6
plot_grid(venn_terms_bp, venn_terms_cc, NULL, NULL, venn_terms_mf, NULL,
          nrow=3, labels=c('a', 'b', '', '', 'c', ''),
          rel_heights = c(1, 0.2, 1))
```

We also evaluate the degree of overlap in enriched CP pathways (when using the raw CP set):

```{r}
venn_cp <- ggvenn(list('HPO' = human_cp$df$ID, 
                       'MGD' = mouse_cp$df$ID,
                       'panUKB' = puk_cp$df$ID),
                   set_name_size = 4, text_size=3, stroke_size=0.3,
                   fill_color=brewer.pal(3, 'Accent'),
                   show_percentage=FALSE)
print(venn_cp)
```

Now we want to test if the overlap is mainly driven by genes that are highly pleiotropic in both datasets. To test for this, we need to construct sets of highly pleiotropic genes that are unique to each data source (trait domain, i.e. we treat HPO and MGD as a single source).

#### Checking with domain-specific genes

```{r}
# Unique gene enrichment
hip_puk_no_mono <- na.omit(merged_data[merged_data$pleiotropy_panUKB == 'high' &
                          is.na(merged_data$HP_high_count) &
                          is.na(merged_data$MP_high_count), 'gene_human'])
hip_hpo_no_puk <- na.omit(merged_data[merged_data$pleiotropy_HPO == 'high' &
                          is.na(merged_data$pleio_degree), 'gene_human'])
hip_mgd_no_puk <- na.omit(merged_data[merged_data$pleiotropy_MGD == 'high' &
                          is.na(merged_data$pleio_degree), 'gene_mouse'])
# HPO unique
hpouq_bp <- make_enrichment_plot(hip_hpo_no_puk, "org.Hs.eg.db", 'BP', 
              'Biological process (HPO unique)', data_source = "POUQ")
hpouq_mf <- make_enrichment_plot(hip_hpo_no_puk, "org.Hs.eg.db", 'MF', 
              'Molecular function (HPO unique)', data_source = "POUQ")
hpouq_cc <- make_enrichment_plot(hip_hpo_no_puk, "org.Hs.eg.db", 'CC', 
              'Cellular component (HPO unique)', data_source = "POUQ")
# Mouse unique
mgduq_bp <- make_enrichment_plot(hip_mgd_no_puk, "org.Mm.eg.db", 'BP', 
              'Biological process (MGD unique)', data_source = "POUQ")
mgduq_cc <- make_enrichment_plot(hip_mgd_no_puk, "org.Mm.eg.db", 'CC', 
              'Cellular component (MGD unique)', data_source = "POUQ")
mgduq_mf <- make_enrichment_plot(hip_mgd_no_puk, "org.Mm.eg.db", 'MF', 
              'Molecular function (MGD unique)', data_source = "POUQ")
# pan-UKB unique
pukuq_bp <- make_enrichment_plot(hip_puk_no_mono, "org.Hs.eg.db", 'BP', 
               'Biological process (pan-UKB unique)', data_source = "POUQ")
pukuq_mf <- make_enrichment_plot(hip_puk_no_mono, "org.Hs.eg.db", 'MF', 
               'Molecular function (pan-UKB unique)', data_source = "POUQ")
pukuq_cc <- make_enrichment_plot(hip_puk_no_mono, "org.Hs.eg.db", 'CC', 
               'Cellular component (pan-UKB unique)', data_source = "POUQ")
venn_terms_bpuq <- ggvenn(list('HPO' = hpouq_bp$df$ID, 
                             'MGD' = mgduq_bp$df$ID,
                             'panUKB' = pukuq_bp$df$ID),
                        set_name_size = 4, text_size=3, stroke_size=0.3,
                        fill_color=brewer.pal(3, 'Accent'),
                        show_percentage=FALSE)
venn_terms_ccuq <- ggvenn(list('HPO' = hpouq_cc$df$ID, 
                             'MGD' = mgduq_cc$df$ID,
                             'panUKB' = pukuq_cc$df$ID),
                        set_name_size = 4, text_size=3, stroke_size=0.3,
                        fill_color=brewer.pal(3, 'Accent'),
                        show_percentage=FALSE)
venn_terms_mfuq <- ggvenn(list('HPO' = hpouq_mf$df$ID, 
                             'MGD' = mgduq_mf$df$ID,
                             'panUKB' = pukuq_mf$df$ID),
                        set_name_size = 4, text_size=3, stroke_size=0.3,
                        fill_color=brewer.pal(3, 'Accent'),
                        show_percentage=FALSE)

# 6 * 6
plot_grid(venn_terms_bpuq, venn_terms_ccuq, 
          NULL, NULL, venn_terms_mfuq, NULL,
          nrow=3, labels=c('a', 'b', '', '', 'c', ''),
          rel_heights = c(1, 0.2, 1))
```

Exporting the updated Extended data figure with term Venns.

```{r}
#| fig-show: hide
# Extended Data Figure 10
plot_grid(venn_terms_bp, venn_terms_bpuq, 
          NULL, NULL,
          venn_terms_cc, venn_terms_ccuq, 
          NULL, NULL,
          venn_terms_mf, venn_terms_mfuq,
          nrow=5, labels=c('a', 'b', '', '', 'c', 'd',
                           '', '', 'e', 'f'),
          rel_heights = c(1, 0.2, 1, 0.2, 1))
```

We can also do the same for the CP pathway set:

```{r}
hpouq_cp <- make_enrichment_plot(hip_hpo_no_puk,  "MSigDB",
              "CP", 'Canonical pathways (human)', use_t2g = TRUE,
              t2g_df=cp_human_t2g, data_source = "POUQ",
                     gs_sizes=gs_sizes)
mgduq_cp <- make_enrichment_plot(hip_mgd_no_puk, "MSigDB",
              "CP", 'Canonical pathways (mouse)', use_t2g = TRUE,
              t2g_df=cp_mouse_t2g, data_source = "POUQ",
                     gs_sizes=gs_sizes)
pukuq_cp <- make_enrichment_plot(hip_puk_no_mono, "MSigDB",
              "CP", 'Canonical pathways (GWAS)', use_t2g = TRUE,
              t2g_df=cp_human_t2g, data_source = "GWASUQ",
                     gs_sizes=gs_sizes)
venn_cp_uq <- ggvenn(list('HPO' = hpouq_cp$df$ID, 
                          'MGD' = mgduq_cp$df$ID,
                          'panUKB' = pukuq_cp$df$ID),
                        set_name_size = 4, text_size=3,
                        stroke_size=0.3, 
                        fill_color=brewer.pal(3, 'Accent'),
                        show_percentage=FALSE)
print(venn_cp_uq)
```

As we can see, the enrichment in pan-UKB data basically disappears (with the exception of a few terms and a slightly higher number of CP pathways). Those that are still detected are, however, distinct from those found in HPO/MGD. This observation indicates the following:

-   the hypothesis that dataset-specific highly pleiotropic genes control the same pathways as the shared ones is not true. It seems that, to the contrary, shared genes control the same set of pathways, and specific genes control specific pathways;

-   There is a likely enrichment of GWAS-specific highly pleiotropic genes with technical causal gene inference artifacts (it explains the overall low rate of enrichment in this set of genes).

Let's also print out the set of pathways that are uniquely enriched in GWAS, and draw the corresponding plot (which may be used in SI):

```{r}
print(pukuq_cp$df$ID)
print(pukuq_cp$plot)
```

I will now combine the plots into the figure panels:

```{r}
fig3_venns <- plot_grid(venn_terms_bp, venn_terms_bpuq,
                        nrow=1, labels=c('a', 'b'),
                        label_size = 20)
fig3_alt_venns <- plot_grid(venn_cp, venn_cp_uq,
                            nrow=1, labels=c('a', 'b'),
                            label_size=20)
```

### Analysis of per-term pleiotropic genes fraction

This analysis aims at looking into how well do the percentages of highly pleiotropic genes according to HPO and pan-UKB correlate within a given GO term. To this end, we define a set of convenience functions (note that the function has the `remove_common_genes` parameter that aims to exclude genes that are highly pleiotropic in both sources, analogous to manipulations for Venn diagrams above).

```{r}
get_share_perc <- function(go_term_id, gotab_df,
                           remove_common_genes) {
  snippet <- gotab_df[gotab_df$GO == go_term_id, ]
  total_uses <- nrow(snippet)
  if (remove_common_genes) {
    snippet <- snippet[!(snippet$pleiotropy_HPO == 'high' &
                         snippet$pleiotropy_panUKB == 'high'), ]
  }
  all_hpo_uses <- sum(snippet$HP_high_count > 0, na.rm=T)
  all_puk_uses <- sum(snippet$pleio_degree > 0, na.rm=T)
  all_any_uses <- sum(snippet$HP_high_count > 0 |
                      snippet$pleio_degree > 0, na.rm=T)
  high_hpo_uses <- sum(snippet$pleiotropy_HPO == 'high', na.rm = T)
  high_puk_uses <- sum(snippet$pleiotropy_panUKB == 'high', na.rm = T)
  return(c(total_uses, c(all_hpo_uses, all_puk_uses, all_any_uses,
           high_hpo_uses, high_puk_uses)/total_uses))
}
make_term_sharing_plot <- function(ontology_type, gotab_df,
              remove_common_genes=FALSE, ax_lim=1) {
  gotab_df <- gotab_df[gotab_df$ONTOLOGY == ontology_type, ]
  print(paste0("Total term count: ", length(unique(gotab_df$GO))))
  term_usage <- table(gotab_df$GO)
  print(paste0('Terms with more than 1 usage : ', sum(term_usage > 1)))
  print(paste0('Terms with more than 10 usages : ', sum(term_usage >= 10)))
  used_terms = names(term_usage[term_usage >= 10])
  gotab_df <- merge(gotab_df, merged_data[, c('gene_human', 
              paste0('pleiotropy_', source_names),
              'pleio_degree', 'HP_high_count')],
              by.x='SYMBOL', by.y='gene_human')
  share_df <- as.data.frame(t(sapply(used_terms, function(x) 
                get_share_perc(x, gotab_df, remove_common_genes))))
  colnames(share_df) <- c('total_uses', 'perc_all_hpo', 'perc_all_panukb', 
                          'perc_any', 'perc_hpo', 'perc_panukb')
  cortest_res <- cor.test(share_df$perc_hpo, 
                          share_df$perc_panukb, 
                          method='spearman')
  print(paste0('Correlation : ', cortest_res$estimate))
  print(paste0('Correlation p-value : ', cortest_res$p.value))
  
  hexpl <- ggplot(share_df, aes(x=perc_hpo, y=perc_panukb)) +
    geom_hex(bins=20, col='black', lwd=0.3) + theme_bw() +
    scale_fill_gradientn(colors=mypal, trans = "log",
                         breaks=c(1, 10, 100, 1000), 
                         name=paste0(ontology_type, ' term count')) +
    xlab('% highly pleiotropic (HPO) ') +
    ylab('% highly pleiotropuc\n (pan-UKB GWAS)') +
    scale_x_continuous(limits=c(0, ax_lim)) +
    scale_y_continuous(limits=c(0, ax_lim)) +
    theme(panel.grid=element_blank())
  
  share_df$ontology_type = ontology_type
  share_df$remove_common = remove_common_genes
  
  ret_val <- list("plot" = hexpl, "df" = share_df, "test.result" = cortest_res)
  
  return(ret_val)
}
```

And then apply them to all gene sets of interest:

```{r}
#| fig-show: hide
bp_sharing <- make_term_sharing_plot("BP", human_gotab)
mf_sharing <- make_term_sharing_plot("MF", human_gotab)
cc_sharing <- make_term_sharing_plot("CC", human_gotab)

bp_sharing_nocomm <- make_term_sharing_plot("BP", human_gotab, 
                       remove_common_genes=TRUE, ax_lim=0.5)
mf_sharing_nocomm <- make_term_sharing_plot("MF", human_gotab, 
                       remove_common_genes=TRUE, ax_lim=0.5)
cc_sharing_nocomm <- make_term_sharing_plot("CC", human_gotab, 
                       remove_common_genes=TRUE, ax_lim=0.5)

all_sharing_df <- rbind(bp_sharing$df, bp_sharing_nocomm$df,
                        mf_sharing$df, mf_sharing_nocomm$df,
                        cc_sharing$df, cc_sharing_nocomm$df)
all_sharing_df$ontology_type = as.factor(all_sharing_df$ontology_type)
all_sharing_df$remove_common = ifelse(all_sharing_df$remove_common,
                      'unique genes', 'all genes')
all_sharing_df$remove_common = as.factor(all_sharing_df$remove_common)

sharing_corrs <- data.frame(ontology_type = rep(c('BP', 'CC', 'MF'), 2),
                    remove_common = rep(c('all genes', 'unique genes'), each=3),
                    perc_hpo = rep(c(0.85, 0.5), each=3),
                    perc_panukb = rep(c(0.9,0.45, 0.65), 2))
sharing_corrs$rho_coeff <- round(c(bp_sharing$test.result$estimate,
                             cc_sharing$test.result$estimate,
                             mf_sharing$test.result$estimate,
                             bp_sharing_nocomm$test.result$estimate,
                             cc_sharing_nocomm$test.result$estimate,
                             mf_sharing_nocomm$test.result$estimate), 2)
```

And then we construct a combined plot of the results with the correlation coefficients in the corners:

```{r}
# Additional Figure (not in text)
go_term_sharing_all <- ggplot(all_sharing_df, aes(x=perc_hpo, y=perc_panukb)) +
  geom_hex(bins=20, col='black', lwd=0.3) + theme_bw() +
  geom_text(data=sharing_corrs, aes(label=rho_coeff), size = 3) +
  facet_grid(vars(ontology_type), vars(remove_common),
             scales = 'free') +
  scale_fill_gradientn(colors=mypal, trans = "log",
                       breaks=c(1, 10, 100, 1000), 
                       name='Term count') +
  xlab('% highly pleiotropic (HPO) ') +
  ylab('% highly pleiotropuc (pan-UKB GWAS)') +
  theme(panel.grid=element_blank(), 
        strip.background = my_ggrect)
print(go_term_sharing_all)
```

And, finally, we look into the top-10 terms with the highest percentage of highly pleiotropic genes across the board (sorted by the minimum percentage across datasets):

```{r}
annotate_gene_percent <- function(enrich_df) {
  enrich_df$k <- sapply(enrich_df$GeneRatio, function(x)
    as.numeric(strsplit(x, '/')[[1]][1]))
  enrich_df$K <- sapply(enrich_df$BgRatio, function(x)
    as.numeric(strsplit(x, '/')[[1]][1]))
  enrich_df$perc_enrich = enrich_df$k/enrich_df$K
  return(enrich_df)
}

make_top10_plot <- function(hpo_df, mgd_df, 
                            gwas_df, 
                            path_size_cut=0) {
  hpo_df <- annotate_gene_percent(hpo_df)
  mgd_df <- annotate_gene_percent(mgd_df)
  gwas_df <- annotate_gene_percent(gwas_df)
  all_df <- merge(merge(hpo_df, mgd_df, by = "ID"),
                  gwas_df, by='ID')
  all_df <- all_df[all_df$K.x >= path_size_cut, c('ID', 'Description', 'perc_enrich.x',
                          'perc_enrich.y', 'perc_enrich')]
  colnames(all_df) = c('ID', 'Description', 'HPO', 'MGD', 'panUKB')
  add_line_break <- function(x, col_lim=40) {
    x <- as.character(x)
    if (str_length(x) > col_lim) {
      x_words <- strsplit(x, ' ')[[1]]
      word_cut = ceiling(length(x_words)/2)
      first_half <- paste(x_words[1:word_cut], collapse=' ')
      second_half <- paste(x_words[(word_cut + 1):length(x_words)],
                           collapse=' ')
      return(paste(first_half, second_half, sep='\n'))
    } else {
      return(x)
    }
  }
  all_df$Description = sapply(all_df$Description, add_line_break)
  all_df$min_perc <- apply(all_df[, 3:5], 1, min)
  all_df <- all_df[order(all_df$min_perc, decreasing = T), ]
  all_df$Description = factor(as.character(all_df$Description),
                          levels=all_df$Description[order(all_df$min_perc)])
  #print(colnames(all_df))
  
  all_df <- melt(head(all_df, 10), id.vars=c('ID', 'Description', 'min_perc'))
  top10_ret_plot <- ggplot(all_df, aes(x=Description, 
                          y=value, fill=variable)) +
    geom_bar(stat='identity', col='black', position='dodge') +
    theme_bw() + theme(panel.grid=element_blank()) +
    coord_flip() + scale_fill_brewer(palette='Accent', name = "Dataset") +
    ylab('Fraction of highly pleiotropic genes') +
    xlab('')
  
  return(top10_ret_plot)
}

path_size_cutoff = 500
shared_top10_bp <- make_top10_plot(human_bp$df, mouse_bp$df,
                            puk_bp$df, path_size_cutoff)
print(shared_top10_bp)
shared_top10_cc <- make_top10_plot(human_cc$df, mouse_cc$df,
                            puk_cc$df, path_size_cutoff)
print(shared_top10_cc)
shared_top10_mf <- make_top10_plot(human_mf$df, mouse_mf$df,
                            puk_mf$df, path_size_cutoff)
print(shared_top10_mf)
```

Now let's make the plots with CPs (both all, with pathway size cutoff, and filtered).

```{r}
shared_top10_cp <- make_top10_plot(human_cp$df, mouse_cp$df,
            puk_cp$df, path_size_cut = 200)
print(shared_top10_cp)

shared_top_indepcp <- make_top10_plot(human_indepcp$df,
                        mouse_indepcp$df, puk_indepcp$df,
                        path_size_cut = 0)
print(shared_top_indepcp)
```

And then we plot a combined figure (as well as an Extended Data Figure) \[also excluded from Qmd output\]:

```{r}
#| fig-show: hide
# Figure 3 [two alt variants] - 6 * 8 in portrait
plot_grid(fig3_venns, shared_top10_bp,
          labels=c('', 'c'), label_size=20, 
          nrow=2, rel_heights = c(0.7, 1))
plot_grid(fig3_alt_venns, shared_top_indepcp,
          labels=c('', 'c'), label_size=20, 
          nrow=2, rel_heights = c(0.7, 1))


# Extended Data Figure 11
plot_grid(shared_top10_bp, shared_top10_cc, 
          shared_top10_mf, nrow=3, 
          labels=c('a', 'b', 'c'), label_size=20)
```

### Searching for god pathway examples

In this section, we try to identify GO biological process terms that have the largest percentage of highly pleiotropic genes in all sources, and a substantial contribution from both Mendelian and complex trait data. This requires a bit of manual curation of the dataframe, the examples that are exported have been taken by examining the `sorted_bp` dataframe.

```{r}
# Trying to identify a good example BP term
sorted_bp <- bp_sharing$df[order(bp_sharing$df$perc_any, 
                                 decreasing = T), ]
export_term_data <- function(term_id) {
  target_genes <- human_gotab[human_gotab$GO == term_id, 'SYMBOL']
  tqrget_data <- merged_data[merged_data$gene_human %in% target_genes,
          c('gene_human', 'pleiotropy_HPO', 
            'pleiotropy_MGD', 'pleiotropy_panUKB',
            'HP_high_count', 'MP_high_count', 'pleio_degree')]
  term_descr <- human_bp$df[human_bp$df$ID == term_id, 'Description']
  write.table(tqrget_data, file=paste0('./pathway_examples/', 
                                       term_id, '_' , term_descr,'.tsv'),
              sep='\t', quote=F, row.names = F)
}
export_term_data('GO:0048009')
export_term_data('GO:0007596')
export_term_data('GO:0002062')
export_term_data('GO:0050679')
export_term_data('GO:0035264')
export_term_data('GO:0051216')
export_term_data('GO:0007507')
export_term_data('GO:0001503')
```

## Analysis of the natural selection signatures

Finally, we turn to the analysis of natural selection effects for genes in different groups of pleiotropy. There are several major sources of selection statisitcs used in the analysis:

-   phastCons scores (downloaded from the UCSC Table Browser);

-   constraint metrics derived from gnomAD (LOEUF and pLI, taken according to gnomAD v.2.1);

-   iHS scores for the British population (taken from the study by Johson et al., 2018, *Nat Ecol Evol*) \[the choice of GBR as the target population was motivated by similarity to the UKB data\];

-   DRC150 (density of recent coalescent events) according to the study of Palamara et al., 2019, *Nat Genet*.

### Evaluating the relationship between positive and negative selection

Before diving into the analysis of negative and positive selection in the context of pleiotropy, let's examine if the metrics of positive and negative selection are in any degree of correlation with each other.

```{r}
# Checking the LOEUF values for genes with higher iHS
merged_data$high_iHS <- (merged_data$IHS_GBR > 4)
ihs_loeuf <- ggplot(merged_data[!(is.na(merged_data$high_iHS)), ], 
       aes(x=high_iHS, y=oe_lof_upper, fill=high_iHS)) +
  geom_violin(scale = "width", col='black') + 
  geom_boxplot(width=0.2, fill='white', col='black') +
  theme_bw() + xlab("iHS > 4") + guides(fill=F) +
  ylab('LOEUF')
merged_data$high_drc <- (merged_data$max_drc_150 > 0.0025)
drc_loeuf <- ggplot(merged_data[!(is.na(merged_data$high_drc)), ], 
       aes(x=high_drc, y=oe_lof_upper, fill=high_drc)) +
  geom_violin(scale = "width", col='black') + 
  geom_boxplot(width=0.2, fill='white', col='black') +
  theme_bw() + xlab("DRC150 > 0.0025") + guides(fill=F) +
  ylab('LOEUF')
# Extended Data Figure 16
plot_grid(ihs_loeuf, drc_loeuf, nrow=1)
aggregate(gene_human~high_iHS+high_drc, merged_data, length)

# Metric correlations - LOEUF, DRC, iHS
ihs_hex <- ggplot(merged_data, aes(x=IHS_GBR, y=oe_lof_upper)) +
  geom_hex(bins=20)
drc_hex <- ggplot(merged_data, aes(x=max_drc_150, y=oe_lof_upper)) +
  geom_hex(bins=20) + scale_x_log10()
plot_grid(ihs_hex, drc_hex, nrow=1)

ggplot(merged_data, aes(x=IHS_GBR, y=max_drc_150)) +
  geom_hex(bins=20) + scale_y_log10()
cor.test(merged_data$max_drc_150, merged_data$IHS_GBR, 
         method='spearman',
         use='complete.obs')
```

### Single metric analysis

#### phastCons

We begin by assessing the individual metrics, starting with the pastCons scores.. First, we look into the gene-level phastCons scores (not CDS-averaged).

```{r}
# Gene-level plots of phastCons, iHS, and DRC150
phastcons_plot <- plot_feature('phast_cons', 'phastCons score', ax_lim = 0.4)
print(phastcons_plot)
phastcons_prop_plot <- plot_feature('phast_cons', 'phastCons score', 
                      cutoff=0.25, cut_name = 'phastCons > 0.25')
print(phastcons_prop_plot)
```

There are few differences, which is expected given the inclusion of non-coding elements. THe same analysis with CDS is clearer:

```{r}
table(is.na(merged_data$phastcons_cds))
phastcons_cds_plot <- plot_feature('phastcons_cds', 'phastCons score (CDS)')
print(phastcons_cds_plot)
phastcons_cds_prop_plot <- plot_feature('phastcons_cds', 'phastCons score (CDS)', 
                                    cutoff=0.6, cut_name = 'phastCons > 0.6')
print(phastcons_cds_prop_plot)
```

And then the same using the locus-wise estimates for 100,000 bp loci around an index SNP.

```{r}
# Locus-level stats (iHS and DRC150) - Figure S9
# Locus-wise phastCons score
phc_locus_plot <- plot_feature('phast_cons', 'phastCons score', 
                gwas_data_file = 'locus_meanphc.tsv')
print(phc_locus_plot)
phc_locus_prop_plot <- plot_feature('phast_cons', 'phastCons score', 
                gwas_data_file = 'locus_meanphc.tsv', 
                cutoff=0.15, cut_name='phastCons > 0.15')
print(phc_locus_prop_plot)
```

#### gnomAD gene-level constraint metrics

We continue with the gnomAD metrics - LOEUF, pLI, and missense variant observed-to-expected ratio. Starting with LOEUF (including additional plots for locus-wise and max-indep data).

```{r}
loeuf_plot <- plot_feature('oe_lof_upper', 'LOEUF')
print(loeuf_plot)

# For maxindep
loeuf_plot_mi <- plot_feature('oe_lof_upper', 'LOEUF', 
                            use_maxindep = TRUE)
print(loeuf_plot_mi)
loeuf_plot_lw <- plot_feature('oe_lof_upper', 'LOEUF', 
                            use_locuswide = TRUE)
print(loeuf_plot_lw)
```

Exporting the SN figures with all gene-level properties \[as usuual, we hide the output from the Qmd\].

```{r}
#| fig-show: hide
# Figure SN2.2
plot_grid(count_plot_lw, ppi_plot_lw, gobp_plot_lw, loeuf_plot_lw,
          nrow=2, labels=c('a', 'b', 'c', 'd'), label_size=20)
# Figure SN3.2
plot_grid(count_plot_mi, ppi_plot_mi, gobp_plot_mi, loeuf_plot_mi,
          nrow=2, labels=c('a', 'b', 'c', 'd'), label_size=20)
```

Next, we consider pLI and missense Z-score. Given the binary tendency of pLI, we also try to do the analysis by splitting genes into tolerant and intolerant.

```{r}
pli_plot <- plot_feature('pLI', 'pLI')
print(pli_plot)
pli_prop_plot <- plot_feature('pLI', 'pLI', cutoff=0.9, cut_name = "pLI > 0.9")
print(pli_prop_plot)
missense_plot <- plot_feature('mis_z', 'Missense z-score')
print(missense_plot)
```

Exporting extended data figure:

```{r}
#| fig-show: hide
# Extended Data Figure 13
plot_grid(pli_plot, missense_plot, nrow=2, labels=c('a', 'b'), label_size=20)
```

#### Positive selection metrics

We begin with iHS values (gene-wise comes first).

```{r}
ihs_plot <- plot_feature('IHS_GBR', 'iHS score')
print(ihs_plot)
ihs_prop_plot <- plot_feature('IHS_GBR', 'iHS score', 
                              cutoff=4, cut_name='iHS > 4')
print(ihs_prop_plot)
# Attempting other iHS cutoffs
plot_feature('IHS_GBR', 'iHS score', cutoff=5, cut_name='iHS > 5')
plot_feature('IHS_GBR', 'iHS score', cutoff=6, cut_name='iHS > 6')
```

And then proceeding to the DRC150 (also in a gene-wise manner):

```{r}
drc_plot <- plot_feature('max_drc_150', 'DRC150 score', log_scale=TRUE)
print(drc_plot)
drc_prop_plot <- plot_feature('max_drc_150', 'DRC150 score', 
                              cutoff=0.0025, cut_name='DRC150 > 0.0025')
print(drc_prop_plot)
```

As Frachon *et al.,* 2017, *Nat Ecol Evol*, as well as some other sources, argue that only a moderate pleiotropy degree should be adaptive, we next check what happens in the high pleiotropy part of the distribution.

```{r}
# High pleiotropy check
sum(merged_data$pleio_degree > 7, na.rm=TRUE)
sum(merged_data[merged_data$pleio_degree > 7, 'max_drc_150'] > 0.0025, na.rm = TRUE)
sum(merged_data$pleio_degree > 10, na.rm=TRUE)
sum(merged_data[merged_data$pleio_degree > 10, 'max_drc_150'] > 0.0025, na.rm = TRUE)
```

And then exporting the final Figure 3 and a corresponding Extended Data Figure:

```{r}
#| fig-show: hide
# Figure 4 --- 8 * 8 (portrait)
plot_grid(phastcons_cds_plot, loeuf_plot, # phastcons_plot, 
          plot_grid(ihs_prop_plot, drc_prop_plot, nrow=1,
                labels=c('c', 'd'), label_size=20),
          nrow=3, labels=c('a', 'b', ''), label_size=20,
          rel_heights = c(0.62, 0.62, 1))

# Extended Data Figure 14
plot_grid(ihs_plot, drc_plot, nrow=2, labels=c('a', 'b'), label_size=24)
```

Finally, we plot the locus-wise iHS and DRC150 values.

```{r}
ihs_locus_plot <- plot_feature('IHS_GBR', 'iHS score', 
                gwas_data_file = 'locus_maxihs.tsv')
print(ihs_locus_plot)
ihs_locus_prop_plot <- plot_feature('IHS_GBR', 'iHS score', 
                gwas_data_file = 'locus_maxihs.tsv', 
                cutoff=4, cut_name='iHS > 4')
print(ihs_locus_prop_plot)

# Locus-wise DRC150 value
drc_locus_plot <- plot_feature('max_drc_150', 'DRC150 score', log_scale=TRUE,
                gwas_data_file = 'locus_maxdrc.tsv')
print(drc_locus_plot)
drc_locus_prop_plot <- plot_feature('max_drc_150', 'DRC150', 
                gwas_data_file = 'locus_maxdrc.tsv', 
                cutoff=0.0025, cut_name='DRC150 > 0.0025')
print(drc_locus_prop_plot)
```

And plotting the extended data figure with the locus-wise positive and negative selection estimates.

```{r}
#| fig-show: hide
# Extended Data Figure 12
plot_grid(phc_locus_plot, ihs_locus_plot, drc_locus_plot, 
          phc_locus_prop_plot, ihs_locus_prop_plot, drc_locus_prop_plot, 
          nrow=2, labels=c('a', 'b', 'c', 'd', 'e', 'f'), label_size=20,
          rel_heights = c(0.7, 1))
```

### Bidirectional selection

As a final step of our analysis pipeline, we look into the overlap of the negative and positive selection signals at highly pleiotropic loci.

```{r}
# Overlapping signatures (based on iHS)
merged_data$bidirectional <- as.numeric(merged_data$IHS_GBR > 4 &
                merged_data$oe_lof_upper < 0.6)
table(merged_data$bidirectional)
sum(merged_data$bidirectional, na.rm=T)/length(na.omit(merged_data$bidirectional))
bidir_ihs <- plot_feature('bidirectional', '', cutoff=0.5, 
                          cut_name='Bidirectional?')

# Overlapping signatures (based on DRC150)
merged_data$bidirectional_b <- as.numeric(merged_data$max_drc_150 > 0.0025 &
                merged_data$oe_lof_upper < 0.6)
table(merged_data$bidirectional_b)
sum(merged_data$bidirectional_b, na.rm=T)/length(na.omit(merged_data$bidirectional_b))
bidir_drc <- plot_feature('bidirectional_b', '', cutoff=0.5, 
                          cut_name='Bidirectional?')

# Extended Data Figure 15
plot_grid(bidir_ihs, bidir_drc, nrow=1, labels=c('a', 'b'), label_size = 20)
```
